---
title: "Smith-Style Tree Simulation and WBE Scaling Workflow"
output:
  html_document:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6.4, fig.height = 4.8, dpi = 150)
pkg_root <- normalizePath("..")
r_files <- list.files(file.path(pkg_root, "R"), full.names = TRUE)
invisible(lapply(r_files, source))
```

## Purpose

This vignette gives a practical end-to-end workflow for the
`scalingPlantArchitecture` package:

- generate branching trees under controlled asymmetry;
- compute shape and scaling diagnostics;
- compare finite-size simulation behavior to analytical symmetric baselines.

The examples are intentionally lightweight so they run quickly during checks.

## Parameterization and reproducibility

```{r}
p <- new_spa_params(
	max_order = 9,
	path_fraction = 0.68,
	seed = 42,
	length_decay = 0.78,
	radius_decay = 0.79,
	branch_prob = 0.98
)

p
```

## Generate a tree

```{r}
tr <- simulate_branching_tree(p)
plot_tree(
	tr,
	preset = "publication",
	highlight_main_path = TRUE,
	title = "Simulated asymmetric branching architecture",
	subtitle = "Branch width is scaled by conduit-radius proxy"
)
```

## Build Smith-style tree panel

```{r, fig.width=11, fig.height=3.8}
plot_smith_tree_panel(
	c(0.29, 0.51, 0.68, 1.00),
	base_params = p,
	preset = "publication",
	show_order_tone = TRUE,
	highlight_main_path = FALSE
)
```

## Core metrics for one tree

```{r}
m <- compute_scaling_metrics(tr)
m
```

## Ensemble scaling diagnostics

```{r}
p_fast <- new_spa_params(max_order = 7, path_fraction = 0.68, seed = 42)
ens <- analyze_ensemble(path_fraction_values = c(0.29, 0.51, 0.68, 1.00), n_rep = 8, base_params = p_fast)
plot_pathfraction_scaling(ens$metrics, preset = "publication", show_ci = TRUE)
ens$fits
```

## Workflow helper for summary tables and baselines

```{r}
exp_out <- run_scaling_experiment(
	path_fraction_values = c(0.29, 0.51, 0.68, 1.00),
	n_rep = 8,
	base_params = p_fast
)

exp_out$summary
head(exp_out$analytical_baseline$table)
```

## Per-tree allometries

```{r, fig.width=10, fig.height=6}
plot_tree_allometries(tr, preset = "publication", show_ci = TRUE)

allom <- compute_tree_allometries(tr)
allom$summary
```

## Path-length frequency distributions

```{r, fig.width=7.5, fig.height=4.5}
plot_path_length_distribution(tr, endpoint = "tips", bins = 20, preset = "publication")
plot_path_length_distribution(tr, endpoint = "all_nodes", bins = 20, preset = "publication")
```

## Symmetric analytical baseline

```{r}
ana <- analytic_symmetric_wbe(max_order = 12)
head(ana$table)
ana$exponent
```

## Interpretation notes

- Finite-depth trees can deviate from asymptotic scaling exponents.
- Different seeds alter realized geometry and may shift fitted slopes.
- Use ensemble summaries (`analyze_ensemble`) for stable comparisons.
- Treat symmetric analytical baselines as reference limits, not guaranteed
	finite-size outcomes.
