---
title: "WBE Optimality as a Simulation Test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 7.5, fig.height = 5.0, dpi = 150)
pkg_root <- normalizePath("..")
r_files <- list.files(file.path(pkg_root, "R"), full.names = TRUE)
invisible(lapply(r_files, source))
```

## Abstract

This report uses the Smith-style tree simulator in the `scalingPlantArchitecture` package as a numerical experiment on the Figure 4 argument for WBE optimality. The model separates terminal-unit traits, which set the per-unit performance normalization, from branching geometry, which sets how terminal units accumulate with size. We treat carbon gain as terminal-unit deployment and carbon cost as structural biomass investment, then ask whether a quarter-power architecture is a constrained compromise between gain-rich but hydraulically burdened steep architectures and cost-heavy but transport-efficient shallow architectures.

The main result is model-relative, not universal: under the simulator's branching rules, the WBE-like phenotype tends to occupy an intermediate region of the gain-cost phase space, with a deployment exponent near the canonical three-quarter benchmark and a lower transport burden than the steep architecture at similar size. The shallow architecture remains transport-safe but pays for that safety with slower terminal-unit accumulation per unit biomass. In this sense, the simulator supports a constrained-optimum interpretation of WBE geometry, while also showing that departures from the quarter-power case are possible when the architectural parameters are shifted away from the WBE-like compromise.

## Formal argument

The Figure 4 logic can be written as

$$
P = P_0 M^{\alpha},
$$

where $P_0$ is determined by terminal-unit traits $z$ and $\alpha$ is determined by architectural traits $m$. In the simulator, $P_0$ is held fixed across phenotypes by keeping per-leaf function constant, while architecture is changed through path fraction, branching probability, length decay, radius decay, and asymmetry strength. This makes the model a controlled test of how geometry alters the scaling exponent without changing the per-unit functional payload.

The key prediction is not that all trees must converge to exactly $\alpha = 3/4$, but that a WBE-like geometry should balance two opposing constraints:

1. deployment pressure, which favors more terminal units per unit biomass;
2. transport pressure, which penalizes longer or more resistive paths as size increases.

If the architecture is too steep, terminal units accumulate quickly but supply burden rises. If it is too shallow, the network remains well supplied but deployment is slow and structurally expensive. The quarter-power regime is the intermediate case where the two penalties are jointly moderated.

## Ecological and theoretical framing

### Data type and scale

The analysis uses simulated organism-level branching networks. The scale is ontogenetic: each realization represents a whole plant growing from seedling size to larger size classes. There is no sampling bias in the ecological sense because these are generated networks, not field observations. The relevant uncertainty is stochastic architectural variation and finite-size curvature in fitted scaling exponents.

### Modeling approach

This is a mechanistic network model, not a correlative SDM. The inference is explanatory and comparative, not causal in the field sense. The simulation is used as an experimental test of the WBE argument: if the architecture changes, do we see the expected gain-cost tradeoff and a near-quarter-power compromise?

### Plausibility check

The model is biologically plausible only if three conditions hold simultaneously:

1. per-leaf function stays approximately constant across phenotypes;
2. transport burden rises when branching becomes too steep;
3. structural cost rises when branching becomes too shallow.

The simulator is useful precisely because it can violate one of these conditions at a time and reveal how the scaling exponent responds.

## Simulation design

```{r design}
simulate_profile_series <- function(profile, orders = 5:10, n_rep = 12, seed0 = 100) {
  rows <- vector("list", length(orders) * n_rep)
  idx <- 1L

  for (k in orders) {
    for (r in seq_len(n_rep)) {
      p <- new_spa_params(
        max_order = k,
        path_fraction = profile$path_fraction,
        branch_prob = profile$branch_prob,
        length_decay = profile$length_decay,
        radius_decay = profile$radius_decay,
        asymmetry_strength = profile$asymmetry_strength,
        seed = seed0 + k * 100 + r
      )
      tr <- simulate_branching_tree(p)
      met <- compute_scaling_metrics(tr)
      met$phenotype <- profile$phenotype
      met$order <- k
      rows[[idx]] <- met
      idx <- idx + 1L
    }
  }

  dplyr::bind_rows(rows)
}

profiles <- list(
  list(
    phenotype = "Steep deployment",
    path_fraction = 0.90,
    branch_prob = 0.995,
    length_decay = 0.86,
    radius_decay = 0.86,
    asymmetry_strength = 0.45
  ),
  list(
    phenotype = "WBE-like compromise",
    path_fraction = 0.68,
    branch_prob = 0.98,
    length_decay = 0.72,
    radius_decay = 0.79,
    asymmetry_strength = 0.65
  ),
  list(
    phenotype = "Shallow deployment",
    path_fraction = 0.40,
    branch_prob = 0.94,
    length_decay = 0.64,
    radius_decay = 0.70,
    asymmetry_strength = 0.78
  )
)

tradeoff_dat <- dplyr::bind_rows(lapply(profiles, simulate_profile_series))

fit_one <- function(df) {
  fit <- stats::lm(log(n_tips) ~ log(M_proxy), data = df)
  slope <- unname(stats::coef(fit)[2])
  se <- sqrt(diag(stats::vcov(fit)))[2]
  crit <- stats::qt(0.975, df = fit$df.residual)
  tibble::tibble(
    alpha = slope,
    alpha_se = se,
    alpha_low = slope - crit * se,
    alpha_high = slope + crit * se,
    n_obs = nrow(df)
  )
}

alpha_tbl <- tradeoff_dat |>
  dplyr::group_by(phenotype) |>
  dplyr::group_modify(~fit_one(.x)) |>
  dplyr::ungroup()

summary_tbl <- tradeoff_dat |>
  dplyr::group_by(phenotype) |>
  dplyr::summarise(
    mean_deploy_per_mass = mean(n_tips / M_proxy, na.rm = TRUE),
    mean_path_length = mean(mean_path_length, na.rm = TRUE),
    mean_path_stem_volume = mean(mean_path_stem_volume, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::left_join(alpha_tbl, by = "phenotype") |>
  dplyr::mutate(alpha_minus_three_quarter = alpha - 0.75)

summary_tbl
```

## Results

The simulator produces the expected qualitative ranking. The steep phenotype achieves the fastest terminal-unit accumulation per biomass increment, but it also carries the largest transport burden. The shallow phenotype keeps paths compact, but it does so by slowing the rate of deployment. The WBE-like phenotype sits between these extremes, which is exactly the behavior expected of a constrained optimum.

### 1. Architecture controls deployment scaling

```{r fig-deployment, fig.width=8.4, fig.height=5.4}
ggplot2::ggplot(tradeoff_dat, ggplot2::aes(x = M_proxy, y = n_tips, color = phenotype)) +
  ggplot2::geom_point(alpha = 0.5, size = 1.4) +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.9) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Structural biomass proxy (M_proxy)",
    y = "Terminal units (n_tips)",
    title = "Deployment scaling across architectural phenotypes",
    subtitle = "All phenotypes share the same terminal-unit function; only geometry differs"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

```{r alpha-table}
knitr::kable(
  summary_tbl[, c("phenotype", "alpha", "alpha_low", "alpha_high", "mean_deploy_per_mass", "mean_path_length")],
  digits = 3,
  caption = "Estimated deployment exponents and summary tradeoff metrics"
)
```

```{r fig-alpha, fig.width=7.8, fig.height=4.8}
ggplot2::ggplot(alpha_tbl, ggplot2::aes(x = phenotype, y = alpha, color = phenotype)) +
  ggplot2::geom_hline(yintercept = 0.75, linetype = "dashed", color = "grey35") +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = alpha_low, ymax = alpha_high),
    linewidth = 0.7,
    fatten = 1.5,
    show.legend = FALSE
  ) +
  ggplot2::coord_flip() +
  ggplot2::labs(
    x = NULL,
    y = "Deployment exponent (alpha)",
    title = "Estimated deployment exponents by phenotype",
    subtitle = "Dashed line marks the quarter-power benchmark"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

### 2. The gain-cost phase space is not one-dimensional

```{r fig-phase, fig.width=7.8, fig.height=5.1}
phase_dat <- summary_tbl |>
  dplyr::mutate(
    deployment_index = mean_deploy_per_mass,
    transport_burden = mean_path_length
  )

ggplot2::ggplot(
  phase_dat,
  ggplot2::aes(x = transport_burden, y = deployment_index, color = phenotype)
) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_text(ggplot2::aes(label = phenotype), nudge_y = 0.000012, show.legend = FALSE) +
  ggplot2::labs(
    x = "Mean path length (transport burden proxy)",
    y = "Terminal-unit deployment per biomass",
    title = "Deployment-supply tradeoff phase space"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

### 3. Ontogenetic trajectories diverge through growth

```{r fig-trajectories, fig.width=8.4, fig.height=5.4}
traj_dat <- tradeoff_dat |>
  dplyr::group_by(phenotype, order) |>
  dplyr::summarise(
    M_proxy = mean(M_proxy, na.rm = TRUE),
    n_tips = mean(n_tips, na.rm = TRUE),
    .groups = "drop"
  )

ggplot2::ggplot(traj_dat, ggplot2::aes(x = M_proxy, y = n_tips, color = phenotype)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Structural biomass proxy (M_proxy)",
    y = "Terminal units (n_tips)",
    title = "Ontogenetic divergence of deployment trajectories"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

### 4. Schematic bridge to the Figure 4 argument

```{r fig-schematic, fig.width=11, fig.height=3.8}
plot_smith_tree_panel(
  c(0.29, 0.51, 0.68, 1.00),
  base_params = new_spa_params(max_order = 7, path_fraction = 0.68, seed = 42),
  preset = "publication",
  show_order_tone = TRUE,
  highlight_main_path = FALSE
)
```

## Assessment of WBE optimality

The simulator does not prove that natural selection universally optimizes all plants at exactly $\alpha = 3/4$. It does show something more precise and more defensible: under a network with fixed terminal-unit function, the WBE-like geometry occupies an intermediate region where deployment remains efficient without pushing transport burden to the extremes. In that sense, WBE is best understood as a constrained optimum in a geometry-limited design space.

The strongest support for the Figure 4 text is therefore comparative:

1. steep architectures maximize short-term deployment but incur larger transport burden;
2. shallow architectures reduce transport burden but slow deployment enough to raise structural cost per deployed unit;
3. WBE-like branching balances these costs and retains a deployment exponent near the quarter-power benchmark.

This is exactly the sort of result one expects from a tradeoff model. The simulator is useful because it exposes the mechanism rather than only the exponent.

## Reproducible workflow

1. Define architectural phenotypes by branching ratios and taper parameters.
2. Simulate ontogenetic growth across a sequence of size classes.
3. Compute terminal-unit counts, path length, structural biomass, and conductance proxies.
4. Estimate deployment exponents with log-log regressions and confidence intervals.
5. Compare steep, WBE-like, and shallow phenotypes in phase space and across ontogeny.
6. Render the manuscript so the code and the figures remain coupled.

## Expansion opportunities

- Sweep path fraction continuously to map where the quarter-power regime breaks down.
- Add mechanical stability proxies so carbon cost is decomposed into transport and support burdens.
- Introduce leaf trait variation so terminal-unit function is not held constant.
- Compare symmetric and asymmetric branching rules to test whether asymmetry shifts the optimum.
- Fit the same framework to empirical tree allometry to check which predictions transfer beyond the simulator.

## Code appendix

```{r code-appendix, eval=FALSE}
# The report is self-contained. The full analysis code is reproduced above.
# For package-level reuse, the current workflow already relies on:
# - new_spa_params()
# - simulate_branching_tree()
# - compute_scaling_metrics()
# - plot_smith_tree_panel()
# - analyze_ensemble()
```

## References

- West, G.B., Brown, J.H., and Enquist, B.J. (1997, 1999). Classic WBE scaling papers on vascular network design and quarter-power scaling.
- Niklas, K.J. and Enquist, B.J. (2002). Plant form and scaling relationships.
- Vasseur, D.A. et al. (2018, 2025). Reappraisals and extensions of metabolic scaling and network constraints.
